Tigth Bounds for Online Balanced Graph Partitioning

### 2020-05-08 paper/disc20clustering.tex
Intro

 \title{Online Balanced Graph Partitioning: Optimizations for Large Clusters}

@@

\begin{abstract}
	Distributed   applications,  including  batch  processing, streaming, scale-out databases,
	or machine learning, generate a significant amount of network traffic. By collocating frequently communicating nodes (e.g., virtual machines) on the same clusters (e.g., server or rack), the network load can be reduced and application performance improved. 
	However, as the communication pattern is a priori unknown and may change over time, it needs to be learned efficiently, in an online manner.
	%
	This paper revisits the online 
	balanced repartitioning problem 
	(introduced by Avin et al.~at DISC 2016)
	which asks for an algorithm that strikes
	an optimal tradeoff between the benefits
	of collocation (i.e., lower network load) 
	and its costs (i.e., migrations). 
	%
	Our first contribution is a significantly improved
	lower bound of $\Omega(k\cdot \ell)$ on the
	competitive ratio, where $\ell$ is the number
	of clusters and $k$ is the cluster size,
	even for a scenario in which the communication
	pattern is static and can be perfectly partitioned;
	we also provide a tight upper bound 
	of $O(k\cdot \ell)$ for this scenario.
	In addition, we present a tight upper bound
	of $\Theta(\ell)$ for $k=3$,
	for the general model in which the
	communication pattern can change arbitrarily
	over time. 
\end{abstract}

@@

\maciek{We study two models. One is a subproblem.}

The general model, where $\sigma$ can be arbitrary
was the original problem introduced
by Avin et al.~\cite{repartition-disc} at DISC 2016.
For simplicity, we will refer to the Perfect Partition
model as the \emph{learning} model (as
one has to learn the static communication graph) 
and to the original as the \emph{repartitioning} model (as one continuously adjusts the partition in response to changing communication graph).

\subsubsection{Partitioning model.}
The \emph{balanced repartitioning} problem (\OBRP{})
is a fundamental learning problem
that finds applications in the context of
distributed systems optimization~\cite{repartition-disc}. We are given a set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

@@

\maciek{The competitive ratio definition and the worst case inputs}

\maciek{Ilustrate the challenges. Rent-or-buy. Decision what to evict}

\subsubsection{Learning model.}
In addition to the general model, we study a simplified Perfect Partition model (also studied by Henzinger et al.~\cite{sigmetrics19_partitioning} (SIGMETRICS 2019)) that we consider for a stronger lower bound.
There, we assume that $\sigma$
reveals the edges of a static graph
that can be perfectly partitioned
(no external requests are required).

@@

\maciek{Offline results}
\maciek{Links other problems: for $k=2$ a flavor of online matching, for $\ell = 2$ it generalizes paging}

@@

\subsection{Our Contributions.}
We provide an improved lower bound 
of $\Omega(k\cdot\ell)$ on the competitive ratio of any online deterministic online algorithm 
even in the learning model;
the best known lower bound so far was $\Omega(k)$,
even for a more general model~\cite{repartition-disc}.
We also present an asymptotically optimal, 
$O(k\cdot \ell)$-competitive algorithm
for the learning model.
For the repartitioning model, we present  
an asymptotically optimal,
$\Theta(\ell)$ algorithm for $k=3$;
the best known upper bound 
so far was $O(\ell^2)$~\cite{repartition-disc}.
Then, we present a strictly $6$-competitive algorithm for $k=2$ that improves upon previous $7$-competitive algorithm with $O(\ell^2)$ additive constant..
In addition, we show a new, improved analysis of the algorithm from \cite{repartition-disc} that depends on the solutions to an interesting combinatorial problem (see Section~\ref{ssec:cascade}, The Cascade Problem).
%
The overview of known results is in Table~\ref{tab:overview}.

\begin{table*}
	\centering
	\renewcommand{\arraystretch}{1.5}
	\begin{tabular}{>{\centering\arraybackslash}p{4.5cm}|>{\centering\arraybackslash}p{4.5cm}>{\centering\arraybackslash}p{4.5cm}}
		\rowcolor{gray!50}
		\textbf{Variant} & \textbf{ Lower bound} &\textbf{Upper bound}\\ \hline 
		\textbf{$k=2$}& 3\hspace{0.3cm}\cite{repartition-disc} & 6\hspace{0.3cm}(\S \ref{sec:k2}) \\ 
		\rowcolor{gray!25}
		\textbf{$k=3$}&  $\Omega(\ell)$ \hspace{0.3cm}(\S \ref{sec:lowerbound})& $O(\ell) $\hspace{0.3cm}(\S \ref{sec:k3})\\
		$k > 3$ & $\Omega(k\cdot \ell)$\hspace{0.3cm}(\S  \ref{sec:lowerbound})&$O(k \cdot \ell \cdot \min \{ k \cdot \ell, \Rep(k, \ell) \})$\hspace{0.1cm} (\S \ref{sec:k3}) \\
		\rowcolor{gray!25}
		Learning model & $\Omega(k\cdot \ell)$\hspace{0.3cm}(\S  \ref{sec:lowerbound})&$O(k \cdot \ell)$\hspace{0.3cm} (\S \ref{sec:ppl}) \\
	\end{tabular}
	\caption{Overview of known results and the results provided in this paper. By $\Rep$ we denote the cost of a single component repartition, and we elaborate on this in Section~\ref{sec:k3}. The results pertain to the general partitioning model, but the last row, that pertains to the learning model for arbitrary $k$ and $\ell$.
	}
	\label{tab:overview}
	\vspace{-7mm}
\end{table*}

@@

>>>>

 \title{Tigth Bounds for Online Balanced Graph Partitioning}

@@

\begin{abstract}
	Distributed   applications,  including  batch  processing, streaming, scale-out databases,
	or machine learning, generate a significant amount of network traffic.
	By collocating frequently communicating nodes (e.g., virtual machines) on the same clusters (e.g., server or rack), we can reduce the network load and  improve application performance. 
	However, the communication pattern is a priori unknown and may change over time, hence it needs to be learned in an online manner.
	%
	This paper revisits the online 
	balanced repartitioning problem 
	(introduced by Avin et al.~at DISC 2016)
	which asks for an algorithm that strikes
	an optimal tradeoff between the benefits
	of collocation (i.e., lower network load) 
	and its costs (i.e., migrations). 
	%
	Our first contribution is a significantly improved
	lower bound of $\Omega(k\cdot \ell)$ on the
	competitive ratio, where $\ell$ is the number
	of clusters and $k$ is the cluster size,
	even for a scenario in which the communication
	pattern is static and can be perfectly partitioned;
	we also provide a tight upper bound 
	of $O(k\cdot \ell)$ for this scenario.
	We contribute a tight upper bound
	of $\Theta(\ell)$ for $k=3$,
	for the general model in which the
	communication pattern can change arbitrarily
	over time.
	We improve the result for $k=2$ by providing a strictly $6$-competitive upper bound for the general model.

\end{abstract}

@@

We study two models in this paper: a general \emph{partitioning} model, and its subproblem, a \emph{learning} variant.
The general partitioning model was introduced by Avin et al.~\cite{repartition-disc} at DISC 2016.
The learning model was introduced by Henzinger et al.~\cite{sigmetrics19_partitioning} at SIGMETRICS 2019.

\noindent
\textbf{Partitioning model.}
The \emph{balanced repartitioning} problem (\OBRP{})
is a fundamental learning problem
that finds applications in the context of
distributed systems optimization~\cite{repartition-disc}. We are given a set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

@@

\subsubsection{Learning model.}

We consider a variant of Online Balanced Partitioning, where communication requests are permanent: once a pair of nodes start communicating, they continue forever.
To avoid infinite cost, each communicating pair must be collocated.
We study instances with finite cost: we assume that the communication graph can be perfectly partitioned in the clusters (no external requests are required).
The algorithm's goal is to \emph{learn} the communication graph, and partition it without too many reconfigurations.

\noindent
\textbf{Adversarial input and the competitive analysis.}
\maciek{The competitive ratio definition and the worst case inputs}

\noindent
\textbf{Challenges.}
The partitioning model has a rent-or-buy dilemas. The decision, what to evict is present in both models.

@@

\maciek{Below is similar to SIDMA, possibly too similar}

The static offline version of~the~partitioning~problem, i.e., a problem variant where
migration is not allowed, where all requests are known in advance, and where
the goal is to find an assignment of $n$ nodes to $\ell$~physical machines, each of~capacity $n/\ell$, is known as the
\emph{$\ell$-balanced graph partitioning problem}. The problem is 
NP-complete, and cannot even be approximated within any finite factor unless P
= NP~\cite{AndRae06}.  The static
variant where $\ell = 2$ corresponds to the minimum bisection problem, which
is already NP-hard~\cite{GaJoSt76}, and 
the currently best approximation ratio is $O(\log n)$~\cite{SarVaz95,ArKaKa99,FeKrNi00,FeiKra02,KraFei06,Raec08}.
The inapproximability of the static variant for general values of $\ell$
motivated research on the bicriteria variant, which can be seen as the offline
counterpart of our capacity augmentation approach. Here, the~goal
is~to~compute a~graph partitioning into $\ell$ components of~size at most~$k$ (where $k > n/\ell$) and the cost of the cut is compared to the optimal (non-augmented)
solution where all components are of a~size at most $n/k$. The variant where
$n \geq 2 \cdot k \cdot \ell$ was considered in
\cite{LeMaTr90,SimTen97,EvNaRS00,EvNaRS99,KrNaSc09}. So far, the~best result~is~an~$O(\!\sqrt{\log n \cdot \log \ell})$-approximation algorithm~\cite{KrNaSc09}.

The studied model is related to online
caching~\cite{SleTar85,FKLMSY91,McGSle91,AcChNo00}, sometimes also referred to
as online caching, where requests for data items (nodes) arrive over time and
need to be served from a cache of finite capacity, and where the number of
cache misses must be minimized. Classic problem variants usually boil down to
finding a smart eviction strategy, such as Least Recently Used (LRU)~\cite{SleTar85}. In our
setting, requests can be served remotely (i.e.,~without fetching the
corresponding nodes to a single physical machine). In this light, our model is more
reminiscent of caching models \emph{with
bypassing}~\cite{EpImLN11,EpImLN15,Irani02}. As a~side result, we show that our problem is
capable of emulating online caching.
A major difference between  these problems is that in the caching problems, each request involves a~single element of the universe, while in our model \emph{both} endpoints of a communication request are subject to~optimization.

@@

\subsection{Our Contributions.}
We provide an improved lower bound 
of $\Omega(k\cdot\ell)$ on the competitive ratio of any online deterministic online algorithm 
even in the learning model;
the best known lower bound so far was $\Omega(k)$,
even for a more general model~\cite{repartition-disc}.
We also present an asymptotically optimal, 
$O(k\cdot \ell)$-competitive algorithm
for the learning model.
For the repartitioning model, we present  
an asymptotically optimal,
$\Theta(\ell)$ algorithm for $k=3$, improving the best known upper bound 
so far $O(\ell^2)$~\cite{repartition-disc}.
Then, we present a strictly $6$-competitive algorithm for $k=2$ that improves upon previous $7$-competitive algorithm with $O(\ell^2)$ additive constant.
%
The overview of known results is in Table~\ref{tab:overview}.

\begin{table*}
	\centering
	\renewcommand{\arraystretch}{1.5}
	\begin{tabular}{>{\centering\arraybackslash}p{4.5cm}|>{\centering\arraybackslash}p{4.5cm}>{\centering\arraybackslash}p{4.5cm}}
		\rowcolor{gray!50}
		\textbf{Variant} & \textbf{ Lower bound} &\textbf{Upper bound}\\ \hline 
		\textbf{$k=2$}& 3\hspace{0.3cm}\cite{repartition-disc} & 6\hspace{0.3cm}(\S \ref{sec:k2}) \\ 
		\rowcolor{gray!25}
		\textbf{$k=3$}&  $\Omega(\ell)$ \hspace{0.3cm}(\S \ref{sec:lowerbound})& $O(\ell) $\hspace{0.3cm}(\S \ref{sec:k3})\\
		$k > 3$ & $\Omega(k\cdot \ell)$\hspace{0.3cm}(\S  \ref{sec:lowerbound})&$O(k^2 \cdot \ell^2)$\hspace{0.1cm} \cite{repartition-disc} \\
		\rowcolor{gray!25}
		Learning model & $\Omega(k\cdot \ell)$\hspace{0.3cm}(\S  \ref{sec:lowerbound})&$O(k \cdot \ell)$\hspace{0.3cm} (\S \ref{sec:ppl}) \\
	\end{tabular}
	\caption{Overview of known results and the results provided in this paper. By $\Rep$ we denote the cost of a single component repartition, and we elaborate on this in Section~\ref{sec:k3}. The results pertain to the general partitioning model, but the last row, that pertains to the learning model for arbitrary $k$ and $\ell$.
	}
	\label{tab:overview}
	\vspace{-7mm}
\end{table*}

@@

\subsection{Warm-up: bounding the reconfiguration for $k=3$}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.3\textwidth]{figs/substitute}
	\caption{Cluster signatures in warm-up k=3 proof}
\end{figure}

\maciek{Rewrite: cluster types should be cluster signatures}
In our analysis, we distinguish among three types of clusters: $C_1, C_2, C_3$. In a cluster of type $C_i$, the size of the largest component contained in this cluster is $i$.
Before bounding the competitive ratio of \TAlg, we introduce the lemma that estimates the cost of a single repartition of \TAlg.

\begin{lemma}
	\label{lem:1req}
	During a single repartition, \TAlg exchanges at most $2$ pairs of nodes.
\end{lemma}

\begin{proof}
	Observe that when the repartition is triggered by \TAlg, the resulting partitioning is component respecting.
	Otherwise, if it does not exist, \TAlg simply ends the phase and performs no repartition.

	%We distinguish between three types of clusters: $C_1, C_2, C_3$,
	%which we define as follows.
	%A cluster of type $C_1$ the cluster contains $3$ singleton components (this is also the initial configuration of any cluster).
	%A cluster of type $C_2$  contains one component of size $2$ and one component of size $1$.
	%Finally, a cluster of type $C_3$  contains one component of size $3$.

	Consider a request between $u$ and $v$ that triggered the repartition and let $U$ and $V$ be their respective clusters.
	Note that $U\neq V$,
	as otherwise, the request would not trigger a~repartitioning.
	We consider cases based on the types of clusters $U$ and $V$.
	Note that this is impossible that either $U$ or $V$ is of type $C_3$, as otherwise, we merge components of size $3$, and no component respecting partitioning exists, a contradiction.
	If either $U$ or $V$ is of type $C_1$, then this cluster can fit the merged component, and the reconfiguration is local within $U$ and $V$.
	For two clusters, any configuration can be reached within two swaps, due to the fact that clusters are indistinguishable.

	Finally, we consider the case where both $U$ and $V$ are of type $C_2$. Note that $(u,v)$ cannot both belong to a component of size $2$, as this would mean that \TAlg has the component of size $4$, a contradiction with the case assumption that the component respecting partitioning exists. 
	Otherwise, if either $u$ or $v$ belongs to a component of size $2$, then it suffices to exchange components of size $1$ between $U$ and $V$.
	Finally, if $u$ and $v$ belong to components of size $1$, then we must place them in a cluster different from $U$ and $V$.
	Note that in such case, a $C_1$-type cluster $W$ exists, as otherwise no component respecting partitioning exists. In this case \TAlg performs one swap, i.e., it exchanges the nodes $u$ and $v$ with any two nodes of $W$.
\end{proof}

Corollary: DET2 is $O(\ell)$-competitive for $k=3$.

\begin{comment}
\section{Bounding the cost of a single reconfiguration of components}

\maciek{Overview of the section, intro. Defer introducing the graph notation, the warm-up does not need it}

\subsection{Characterization of the migration graph}

Consider a component merge action performed by \TAlg that triggers a reconfiguration from a configuration $C_I$ to a configuration $C_F$.
To execute the reconfiguration, the nodes of two non-collocated components migrate to a common cluster.
Each cluster contains exactly $k$ nodes, thus other nodes migrate to take the place of nodes of merged components.
Their place must be occupied, too, and this way a single merge action may trigger migrations in clusters not directly involved in a merge action (for an example, see Section~\ref{ssec:example}). 
Recall that both $C_I$ and $C_F$ are component-respecting configurations, hence a migration of a single node entails migrations of all nodes of its cluster.

In Section~\ref{ssec:cascade}, we upper bound the cost incurred by a reconfiguration that follows a single component merge.
To this end, we introduce a \emph{migration graph} that models the reconfiguration.
In the following, we characterize the structure of the migration graph.

\noindent
\textbf{Vertices of the migration graph and the core vertices.}
The vertices of the migration graph are all $\ell$ clusters of the instance.
We distinguish the \emph{core vertices} that correspond to clusters directly involved in the merge operation: the clusters containing the to-be-merged components in $C_I$ and the cluster containing the merged component in $C_F$.
There are at most $3$ core vertices in each migration graph (at most two components participate in the merge, and these may migrate to a third cluster).

\noindent
\textbf{Edges and their labels.}
The edges of the migration graph denote the migration of nodes from $C_I$ to $C_F$.
Each edge is labeled with a number $m$ that denotes the number of migrated nodes between clusters.
Each edge is directed from the cluster that contained the nodes to the cluster they migrated to.
Both edges might exists between a pair of nodes, as whole components of nodes migrate, and the back-and-forth exchange may be needed for the configuration $C_F$ to be component-respecting.

\noindent
\textbf{Vertex degree.}
At most $k$ nodes may migrate from a cluster and at most $k$ nodes may migrate into a cluster.
Thus, the sum of the labels for both ingoing and outgoing edges of each vertex is at most $k$.
This implies that the number of ingoing and outgoing edges is also bounded by $k$.

\noindent
\textbf{Flow preservation.}
In any feasible configuration, each cluster contains exactly $k$ nodes.
Thus, for any vertex, the sum of labels of ingoing edges must equal the sum of labels of outgoing edges.
\maciek{Can we say that this implies that the graph is a circulation?}

\noindent
\textbf{Migration graph and the cost of reconfiguration.}
The cost of cluster reconfiguration equals the number of exchanged nodes multiplied by $\alpha$.
From the standpoint of a migration graph, this corresponds to the sum of labels on edges of the graph.
A trivial upper bound on the cost is the size of the connected component in the migration graph multiplied by $2\alpha k$.

Note that for a single component merge, multiple migration graphs may exist, and multiple of them might have the optimal cost.
In Section \ref{ssec:cascade}, we show that among all migration graphs for given component merge, there exists an optimal graph with the bounded cycle length.
Combined with other properties, this allows to bound the size of the migration graph, and consequently bound the cost of each repartition, that results in a bound of the competitive ratio for \TAlg.

\subsection{Costly cluster reconfiguration example}
\label{ssec:example}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.3\textwidth]{figs/substitute}
	\caption{A costly cascade example. On the left: the configuration prior to the merge and the merge of two components (denoted solid). On the right: corresponding migration graph for the optimal (unique) reconfiguration.}
\end{figure}

In this section we present an example of a merge action that results in a $k/3$(?) length cycle in an optimal migration graph.

\subsection{Bounds on the size of a migration graph}
\label{ssec:cascade}

In \cite{repartition-disc}, authors trivially bounded the reconfiguration cost for each merge action by $k \cdot \ell$.
This roughly corresponds to migrating every node in every cluster.
Now, we bound the cost of a single repartition of \TAlg as a function of $k$.
Note that the bound of $k \cdot \ell$ is still valid, and the resulting bound is the minimum of these two.

We begin, by bounding the length of each cycle in one of optimal solutions.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.3\textwidth]{figs/substitute}
	\caption{Illustration to bounding the cycle length.}
	\label{fig:cascade-illustration}
\end{figure}

\begin{theorem}
	\label{th:cascade-cycles}
	Consider a single merge action of \TAlg.
	There exists an optimal cost reconfiguration with the migration graph, where each cycle have length at most $k$.
\end{theorem}

\begin{proof}
	Assume the contrary, i.e., for each optimal cluster reconfiguration, its migration graph contains a cycle at least $k+1$.

	Fix a reconfiguration and its migration graph $G$ that has the shortest length $L$ of the longest cycle, and has the least number of cycles of length $L$.
	Let $p = \lbrace v_1, v_2, \ldots, v_L \rbrace$ be any cycle of length $L$.
	As edge labels are from $\{ 1, 2, \ldots, k \}$, there exists a pair of edges $e, f$ with equal labels.

	We show that in such case, we can modify the solution without affecting its cost and feasibility to decrease the number of cycles of length $L$ by $1$.
	This would contradict the assumption that we consider a solution with the minimal number of cycles of length $L$.

	We construct a new migration graph $G'$ of equal cost by swapping the (destination) endpoints of $e$ and $f$ (cf. Figure~\ref{fig:cascade-illustration})
	Formally, let $\lbrace a, b \rbrace = e$ and $\lbrace c, d \rbrace = f$.
	We remove $e$ and $f$ and add $e' = \lbrace a, d \rbrace$ and $f' = \lbrace c, d \rbrace$.
	This way we produced a feasible solution.

	It remains to show that while decreasing the number of cycles of length $L$, we did not created any new cycles of length at least $L$.
	Assume that adding $e'$ created a new cycle of length at least $L$.
	However, $a$ and $d$ were already connected in $G$ by a directed path of length at least $2$, and a cycle strictly longer than $L$ existed in $G$.

	Assume that adding $f'$ created a new cycle $s^*$ of length $L' \geq L$.
	Contrary to the previous case, the directed cycle from $e$ to $b$ might not exist in $G$.
	However, this would mean that $b$ and $e$ were connected by a path of length $L'-1$.
	This means that we might have followed this alternative path from $b$ to $e$ to obtain a cycle strictly longer than $L$ (cf. Figure~\ref{fig:cascade-illustration}).

\begin{theorem}
	Consider a single repartition of DET2.
	The number of clusters it involves is bounded by $k^k$.
	\label{th:cascade}
\end{theorem}

\begin{proof}
	Each node has degree at most $k$.
	Every subset has flow conservation property, hence each cycle traverses one of the nodes of graph origin/core. (no subsets with just incoming edges to the origin part)
	Combine with Theorem \ref{th:cascade-cycles}.

Corollary: Algorithm DET2 is $O(\ell \cdot k \cdot \min \{ \ell, k^k \})$-competitive.
\end{comment}
